---
title: "DE Analysis: 11-sample subset"
author: "Sam Remmey"
date: "2025-02-07"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## DE analysis on 8 POAG samples vs 3 control samples

```{r}
gene <- data.frame(readxl::read_excel("gene_count_16.xls"))
```


```{r}
head(gene)
```

```{r}
#install.packages("BiocManager")  # If not already installed
#BiocManager::install("DESeq2")

library(tidyverse)
library(DESeq2)
library(ggplot2)
library(pheatmap)
```


### Running DE on 11 samples (8 POAG, 3 cadaver controls):
```{r}
counts_matrix <- gene %>%
  select(gene_id, OS0658, OS21014, OD21014, AD24, AD25, RNA41, RNA45, RNA47, RNA48, RNA53, RNA44)
rownames(counts_matrix) <- counts_matrix$gene_id
counts_matrix <- counts_matrix %>%
  select(-gene_id)
counts_matrix <- as.matrix(counts_matrix)  # Convert to numeric


samples <- c('OS0658','OS21014','OD21014','AD24','AD25','RNA41','RNA45','RNA47','RNA48','RNA53','RNA44')
label <- c('CTRL','CTRL','CTRL','POAG','POAG','POAG','POAG','POAG','POAG','POAG','POAG')
labels <- data.frame(samples, label)
rownames(labels) <- labels$samples
labels <- labels %>%
  select(-samples)
```

```{r}
 ## Create a DESeqDataSet object
dds <- DESeqDataSetFromMatrix(countData = counts_matrix,
                              colData = labels,
                              design = ~ label)  # 'condition' should be a column in sample_info

# Filter low-count genes (optional)
#dds <- dds[rowSums(counts(dds)) > 10, ]

## Run DESeq analysis
dds <- DESeq(dds)
res <- results(dds)
```


```{r}
# p values
res <- res[order(res$padj), ]
head(res)

# Filter for Significant DEGs (Adjusted p-value < 0.05, Log2FC > 1)
sig_genes_11 <- subset(res, padj < 0.05 & abs(log2FoldChange) > 1)
```

```{r}
# volcano plot
ggplot(as.data.frame(res), aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(alpha = 0.5) +
  geom_point(data = sig_genes_11, aes(x = log2FoldChange, y = -log10(padj)), color = "red") +
  theme_minimal() +
  xlab("Log2 Fold Change") +
  ylab("-Log10 Adjusted P-Value") +
  ggtitle("Volcano Plot of Differentially Expressed Genes")

```

```{r}
# Heatmap of Top Genes
top_genes_11 <- rownames(sig_genes_11)[1:50]  # Select top 50 genes
all_genes_11 <- rownames(sig_genes_11)
normalized_counts <- counts(dds, normalized = TRUE)
pheatmap(normalized_counts[top_genes_11, ])
```

```{r}
vsd_11 <- vst(dds, blind = FALSE)
# PCA
plotPCA(vsd_11, intgroup = "label") +
  theme_minimal() +
  ggtitle("PCA of RNA-seq Data: 11 Samples")
```



```{r}
top_genes_11
```


## Compare to DE analysis results from 6-sample subset

Look at overlap between genes differentially expressed across 8 POAG (bad samples)/3 cadaver and 3 POAG (good samples)/3 cadaver:
```{r}
# all_genes_6 object here comes from DE_6.Rmd file

# **define the all_genes_6 variable after defining sig_genes_6 in the DE_6.Rmd file:**
# all_genes_6 <- rownames(sig_genes_6)

overlap <- intersect(all_genes_6, all_genes_11)
overlap
```


Functional enrichment analysis:

```{r}
R.version.string
install.packages(c("httr", "jsonlite", "curl"))

```

```{r}
install.packages("gprofiler2")

```
```{r}
library(gprofiler2)
```

```{r}
# Run functional enrichment analysis
gost_results <- gost(query = overlap, 
                     organism = "hsapiens",  # Human genes
                     sources = c("GO:BP", "GO:MF", "GO:CC", "KEGG", "REACTOME"), 
                     significant = TRUE)

# Print results
head(gost_results$result)

```

```{r}
# Extract top 10 enriched terms
top_terms <- gost_results$result[1:10,]

ggplot(top_terms, aes(x=reorder(term_name, -p_value), y=-log10(p_value))) +
  geom_bar(stat="identity", fill="steelblue") +
  coord_flip() +
  labs(title="Top Enriched Terms", x="Pathway/GO Term", y="-log10(p-value)") +
  theme_minimal()

```

```{r}
# Interactive manhattan plot
gostplot(gost_results, capped = TRUE, interactive = TRUE)
```






Note:

Only one gene was identified as being differentially expressed across the POAG and NTG samples. 

"Mutations in this gene have been associated with a number of endothelial corneal dystrophies including recessive corneal endothelial dystrophy 2, corneal dystrophy and perceptive deafness, and Fuchs endothelial corneal dystrophy" (https://www.genecards.org/cgi-bin/carddisp.pl?gene=SLC4A11&keywords=ENSG00000088836). 



