---
title: "DE Analysis"
author: "Sam Remmey"
date: "2025-02-07"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}
gene <- data.frame(readxl::read_excel("gene_count_16.xls"))
```


```{r}
head(gene)
```

```{r}
#install.packages("BiocManager")  # If not already installed
#BiocManager::install("DESeq2")

library(DESeq2)
library(ggplot2)
library(pheatmap)
```

```{text}
### Framework for conducting DE:

## Load count matrix and metadata
counts_matrix <- read.csv("gene_counts.csv", row.names = 1)
sample_info <- read.csv("sample_metadata.csv")

# Convert count matrix to numeric; need this to be a counts matrix with genes as rows, samples as columns and raw read counts as values
counts_matrix <- as.matrix(counts_matrix)


## Create a DESeqDataSet object
dds <- DESeqDataSetFromMatrix(countData = counts_matrix,
                              colData = sample_info,
                              design = ~ condition)  # 'condition' should be a column in sample_info

# Filter low-count genes (optional)
dds <- dds[rowSums(counts(dds)) > 10, ]


## Run DESeq analysis
dds <- DESeq(dds)
res <- results(dds)


## Explore results

# p values
res <- res[order(res$padj), ]
head(res)

# Filter for Significant DEGs (Adjusted p-value < 0.05, Log2FC > 1)
sig_genes <- subset(res, padj < 0.05 & abs(log2FoldChange) > 1)

# volcano plot
ggplot(as.data.frame(res), aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(alpha = 0.5) +
  geom_point(data = sig_genes, aes(x = log2FoldChange, y = -log10(padj)), color = "red") +
  theme_minimal() +
  xlab("Log2 Fold Change") +
  ylab("-Log10 Adjusted P-Value") +
  ggtitle("Volcano Plot of Differentially Expressed Genes")

# Heatmap of Top Genes
top_genes <- rownames(sig_genes)[1:50]  # Select top 50 genes
normalized_counts <- counts(dds, normalized = TRUE)
pheatmap(normalized_counts[top_genes, ])


## Save results
#write.csv(as.data.frame(res), "DEG_results.csv")

```



### Running DE on 10 samples (8 POAG, 2 NTG):
```{r}
library(dplyr)

counts_matrix <- gene %>%
  select(gene_id, AD24, AD25, RNA41, RNA45, RNA47, RNA48, RNA53, RNA44, RNA54, RNA61)
rownames(counts_matrix) <- counts_matrix$gene_id
counts_matrix <- counts_matrix %>%
  select(-gene_id)
counts_matrix <- as.matrix(counts_matrix)  # Convert to numeric


samples <- c('AD24','AD25','RNA41','RNA45','RNA47','RNA48','RNA53','RNA44','RNA54','RNA61')
label <- c('POAG','POAG','POAG','POAG','POAG','POAG','POAG','POAG','NTG','NTG')
labels <- data.frame(samples, label)
rownames(labels) <- labels$samples
labels <- labels %>%
  select(-samples)
```

```{r}
## Create a DESeqDataSet object
dds <- DESeqDataSetFromMatrix(countData = counts_matrix,
                              colData = labels,
                              design = ~ label)  # 'condition' should be a column in sample_info

# Filter low-count genes (optional)
#dds <- dds[rowSums(counts(dds)) > 10, ]

## Run DESeq analysis
dds <- DESeq(dds)
res <- results(dds)
```


```{r}
# p values
res <- res[order(res$padj), ]
head(res)

# Filter for Significant DEGs (Adjusted p-value < 0.05, Log2FC > 1)
sig_genes <- subset(res, padj < 0.05 & abs(log2FoldChange) > 1)
```

```{r}
# volcano plot
ggplot(as.data.frame(res), aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(alpha = 0.5) +
  geom_point(data = sig_genes, aes(x = log2FoldChange, y = -log10(padj)), color = "red") +
  theme_minimal() +
  xlab("Log2 Fold Change") +
  ylab("-Log10 Adjusted P-Value") +
  ggtitle("Volcano Plot of Differentially Expressed Genes")

```

```{r}
# Heatmap of Top Genes
top_genes <- rownames(sig_genes)[1:50]  # Select top 50 genes
normalized_counts <- counts(dds, normalized = TRUE)
#pheatmap(normalized_counts[top_genes, ])
```

```{r}
sig_genes
```

Only one gene was identified as being differentially expressed across the POAG and NTG samples. 

"Mutations in this gene have been associated with a number of endothelial corneal dystrophies including recessive corneal endothelial dystrophy 2, corneal dystrophy and perceptive deafness, and Fuchs endothelial corneal dystrophy" (https://www.genecards.org/cgi-bin/carddisp.pl?gene=SLC4A11&keywords=ENSG00000088836). 



