---
title: "DE Full Report"
author: "Sam Remmey, James Siegener, Jacob Turner"
date: "2025-02-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("BiocManager")  # If not already installed
#BiocManager::install("DESeq2")

library(tidyverse)
library(DESeq2)
library(ggplot2)
library(pheatmap)
```

## Differential Expression Analysis

We've conducted DE analyses on both the high confidence 6-sample subset (3 POAG, 3 cadaver) as well as an 11-sample subset we formed by combining the 3 high confidence cadaver control samples with the 8 low confidence POAG samples. This report contains results of both series of analyses as well as a list of extracted genes that were found to be differentially expressed in both subsets, followed by functional enrichment analysis conducted on this extracted set of genes. 


### DE: 6 samples

```{r}
gene_6 <- readxl::read_excel("gene_count_6.xls")

counts_6 <- gene_6 %>% select(gene_id, OS0658, OS21014, OD21014, AD22, RNA35, RNA62)

counts_6 <- counts_6 %>% column_to_rownames(var = "gene_id")

counts_6 <- as.matrix(counts_6)

mode(counts_6) <- "numeric"

sample_info_6 <- data.frame(
    row.names = colnames(counts),   # Sample names
    condition = c("control", "control", "control", "glaucoma", "glaucoma", "glaucoma")  # Group labels
)

sample_info_6$condition <- factor(sample_info_6$condition, levels = c("control", "glaucoma"))
```

```{r}
dds_6 <- DESeqDataSetFromMatrix(countData = counts_6,
                              colData = sample_info_6,
                              design = ~ condition)

# Filter out low-expression genes
# dds_6 <- dds_6[rowSums(counts_6(dds_6)) > 10, ]

# Run DESeq2 normalization and differential expression analysis
dds_6 <- DESeq(dds_6)

# Extract results (glaucoma vs. control)
res_6 <- results(dds_6, contrast = c("condition", "glaucoma", "control"))

# View summary
summary(res_6)
```

```{r}
sig_genes_6 <- subset(res_6, padj < 0.05 & abs(log2FoldChange) > 1)
all_genes_6 <- rownames(sig_genes_6)

head(res_6)
```

```{r}
# Volcano plot
res_df_6 <- as.data.frame(res_6) %>%
  rownames_to_column(var = "gene") %>%   
  drop_na()  

res_df_6$log2FoldChange <- as.numeric(res_df_6$log2FoldChange)
res_df_6$padj <- as.numeric(res_df_6$padj)

ggplot(res_df_6, aes(x = log2FoldChange, y = -log10(padj), color = padj < 0.05)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = c("black", "red")) +  # Red for significant genes
  theme_minimal() +
  labs(title = "Volcano Plot", x = "Log2 Fold Change", y = "-Log10 Adjusted P-value")

```

```{r}
# Extract normalized counts
vsd_6 <- vst(dds_6, blind = FALSE)
top_genes_6 <- rownames(sig_genes_6)[1:25]

# Heat map
pheatmap(assay(vsd_6)[top_genes_6, ], cluster_rows = TRUE, cluster_cols = TRUE)
```

```{r}
# PCA
plotPCA(vsd_6, intgroup = "condition") +
  theme_minimal() +
  ggtitle("PCA of RNA-seq Data: 6 Samples")
```


### DE: 11 samples

```{r}
gene_16 <- data.frame(readxl::read_excel("gene_count_16.xls"))

counts_matrix <- gene_16 %>%
  select(gene_id, OS0658, OS21014, OD21014, AD24, AD25, RNA41, RNA45, RNA47, RNA48, RNA53, RNA44)
rownames(counts_matrix) <- counts_matrix$gene_id
counts_matrix <- counts_matrix %>%
  select(-gene_id)
counts_matrix <- as.matrix(counts_matrix)  # Convert to numeric


samples <- c('OS0658','OS21014','OD21014','AD24','AD25','RNA41','RNA45','RNA47','RNA48','RNA53','RNA44')
label <- c('CTRL','CTRL','CTRL','POAG','POAG','POAG','POAG','POAG','POAG','POAG','POAG')
labels <- data.frame(samples, label)
rownames(labels) <- labels$samples
labels <- labels %>%
  select(-samples)
```

```{r}
 ## Create a DESeqDataSet object
dds <- DESeqDataSetFromMatrix(countData = counts_matrix,
                              colData = labels,
                              design = ~ label)  # 'condition' should be a column in sample_info

# Filter low-count genes (optional)
#dds <- dds[rowSums(counts(dds)) > 10, ]

## Run DESeq analysis
dds <- DESeq(dds)
res <- results(dds)
```

```{r}
# p values
res <- res[order(res$padj), ]
head(res)

# Filter for Significant DEGs (Adjusted p-value < 0.05, Log2FC > 1)
sig_genes_11 <- subset(res, padj < 0.05 & abs(log2FoldChange) > 1)
```

```{r}
# Volcano plot
ggplot(as.data.frame(res), aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(alpha = 0.5) +
  geom_point(data = sig_genes_11, aes(x = log2FoldChange, y = -log10(padj)), color = "red") +
  theme_minimal() +
  xlab("Log2 Fold Change") +
  ylab("-Log10 Adjusted P-Value") +
  ggtitle("Volcano Plot of Differentially Expressed Genes")

```

```{r}
# Heatmap of Top Genes
top_genes_11 <- rownames(sig_genes_11)[1:25]  # Select top 25 genes
all_genes_11 <- rownames(sig_genes_11)
normalized_counts <- counts(dds, normalized = TRUE)
pheatmap(normalized_counts[top_genes_11, ])
```

```{r}
# PCA

vsd_11 <- vst(dds, blind = FALSE)

plotPCA(vsd_11, intgroup = "label") +
  theme_minimal() +
  ggtitle("PCA of RNA-seq Data: 11 Samples")
```


## Comparison

#### Identify overlap in differentially expressed genes. 

```{r}
overlap <- intersect(all_genes_6, all_genes_11)
length(overlap)
```


## Functional Enrichment Analysis

```{r}
#install.packages(c("httr", "jsonlite", "curl"))
#install.packages("gprofiler2")
library(gprofiler2)
```

```{r}
# Run functional enrichment analysis
gost_results <- gost(query = overlap, 
                     organism = "hsapiens",  # Human genes
                     sources = c("GO:BP", "GO:MF", "GO:CC", "KEGG", "REACTOME"), 
                     significant = TRUE)

# Print results
head(gost_results$result)
```

```{r}
# Extract top 10 enriched terms
top_terms <- gost_results$result[1:10,]

ggplot(top_terms, aes(x=reorder(term_name, -p_value), y=-log10(p_value))) +
  geom_bar(stat="identity", fill="steelblue") +
  coord_flip() +
  labs(title="Top Enriched Terms", x="Pathway/GO Term", y="-log10(p-value)") +
  theme_minimal()

```

```{r}
# Interactive manhattan plot
gostplot(gost_results, capped = TRUE, interactive = TRUE)
```




