---
title: "DE Full Report"
author: "Sam Remmey, James Siegener, Jacob Turner"
date: "2025-02-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("BiocManager")  # If not already installed
#BiocManager::install("DESeq2")

library(tidyverse)
library(DESeq2)
library(ggplot2)
library(pheatmap)
```

## Differential Expression Analysis

We've conducted DE analyses on both the high confidence 6-sample subset (3 POAG, 3 cadaver) as well as an 11-sample subset we formed by combining the 3 high confidence cadaver control samples with the 8 low confidence POAG samples. This report contains results of both series of analyses as well as a list of extracted genes that were found to be differentially expressed in both subsets, followed by functional enrichment analysis conducted on this extracted set of genes. 


Global variables:
```{r}
p_val <- 5e-10
```


### DE: 6 samples

```{r}
gene_6 <- readxl::read_excel("gene_count_6.xls")

counts_6 <- gene_6 %>% select(gene_id, OS0658, OS21014, OD21014, AD22, RNA35, RNA62)

counts_6 <- counts_6 %>% column_to_rownames(var = "gene_id")

counts_6 <- as.matrix(counts_6)

mode(counts_6) <- "numeric"

sample_info_6 <- data.frame(
    row.names = colnames(counts),   # Sample names
    condition = c("control", "control", "control", "glaucoma", "glaucoma", "glaucoma")  # Group labels
)

sample_info_6$condition <- factor(sample_info_6$condition, levels = c("control", "glaucoma"))
```

```{r}
dds_6 <- DESeqDataSetFromMatrix(countData = counts_6,
                              colData = sample_info_6,
                              design = ~ condition)

# Filter out low-expression genes
# dds_6 <- dds_6[rowSums(counts_6(dds_6)) > 10, ]

# Run DESeq2 normalization and differential expression analysis
dds_6 <- DESeq(dds_6)

# Extract results (glaucoma vs. control)
res_6 <- results(dds_6, contrast = c("condition", "glaucoma", "control"))

# View summary
summary(res_6)
```

```{r}
## Extract significant genes
sig_genes_6_adj <- subset(res_6, pvalue < p_val & abs(log2FoldChange) > 1)
all_genes_6_adj <- rownames(sig_genes_6_adj)
length(all_genes_6_adj)
```

```{r}
# Volcano plot
res_df_6 <- as.data.frame(res_6) %>%
  rownames_to_column(var = "gene") %>%   
  drop_na()  

res_df_6$log2FoldChange <- as.numeric(res_df_6$log2FoldChange)
res_df_6$padj <- as.numeric(res_df_6$padj)
res_df_6$pvalue <- as.numeric(res_df_6$pvalue)

ggplot(res_df_6, aes(x = log2FoldChange, y = -log10(padj), color = pvalue < p_val)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = c("black", "red")) +  # Red for significant genes
  theme_minimal() +
  labs(title = "Volcano Plot", x = "Log2 Fold Change", y = "-Log10 Adjusted P-value")

```

```{r}
# Extract normalized counts
vsd_6 <- vst(dds_6, blind = FALSE)
top_genes_6 <- rownames(sig_genes_6_adj)[1:10] # top 10

# Heat map
pheatmap(assay(vsd_6)[top_genes_6, ], cluster_rows = TRUE, cluster_cols = TRUE)
```

```{r}
# PCA
plotPCA(vsd_6, intgroup = "condition") +
  theme_minimal() +
  ggtitle("PCA of RNA-seq Data: 6 Samples")
```


### DE: 11 samples

```{r}
gene_16 <- data.frame(readxl::read_excel("gene_count_16.xls"))

counts_matrix <- gene_16 %>%
  select(gene_id, OS0658, OS21014, OD21014, AD24, AD25, RNA41, RNA45, RNA47, RNA48, RNA53, RNA44)
rownames(counts_matrix) <- counts_matrix$gene_id
counts_matrix <- counts_matrix %>%
  select(-gene_id)
counts_matrix <- as.matrix(counts_matrix)  # Convert to numeric


samples <- c('OS0658','OS21014','OD21014','AD24','AD25','RNA41','RNA45','RNA47','RNA48','RNA53','RNA44')
label <- c('CTRL','CTRL','CTRL','POAG','POAG','POAG','POAG','POAG','POAG','POAG','POAG')
labels <- data.frame(samples, label)
rownames(labels) <- labels$samples
labels <- labels %>%
  select(-samples)
```

```{r}
 ## Create a DESeqDataSet object
dds <- DESeqDataSetFromMatrix(countData = counts_matrix,
                              colData = labels,
                              design = ~ label)

# Filter low-count genes (optional)
#dds <- dds[rowSums(counts(dds)) > 10, ]

## Run DESeq analysis
dds <- DESeq(dds)
res <- results(dds)
```

```{r}
# p values
res <- res[order(res$padj), ]
head(res)

# Extract significant genes
sig_genes_11_adj <- subset(res, pvalue < p_val & abs(log2FoldChange) > 1)
all_genes_11_adj <- rownames(sig_genes_11_adj)
```

```{r}
# Compare number of significant genes from 11-sample vs 6-sample
length(all_genes_11_adj)
length(all_genes_6_adj)
```



```{r}
# Volcano plot
res_df_11 <- as.data.frame(res) %>%
  rownames_to_column(var = "gene") %>%   
  drop_na()  

res_df_11$log2FoldChange <- as.numeric(res_df_11$log2FoldChange)
res_df_11$padj <- as.numeric(res_df_11$padj)
res_df_11$pvalue <- as.numeric(res_df_11$pvalue)

ggplot(res_df_11, aes(x = log2FoldChange, y = -log10(padj), color = pvalue < p_val)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = c("black", "red")) +  # Red for significant genes
  theme_minimal() +
  labs(title = "Volcano Plot", x = "Log2 Fold Change", y = "-Log10 Adjusted P-value")

```

```{r}
# Heatmap of Top Genes
top_genes_11 <- rownames(sig_genes_11_adj)[1:25]  # top 25 genes
all_genes_11 <- rownames(sig_genes_11_adj)
normalized_counts <- counts(dds, normalized = TRUE)
pheatmap(normalized_counts[top_genes_11, ])
```

```{r}
# PCA

vsd_11 <- vst(dds, blind = FALSE)

plotPCA(vsd_11, intgroup = "label") +
  theme_minimal() +
  ggtitle("PCA of RNA-seq Data: 11 Samples")
```


PCA again: this time add back the 3 high confidence POAG samples and run PCA on that.
```{r}
gene_16 <- data.frame(readxl::read_excel("gene_count_16.xls"))

counts_matrix2 <- gene_16 %>%
  select(gene_id, OS0658, OS21014, OD21014, AD22, RNA35, RNA62, AD24, AD25, RNA41, RNA45, RNA47, RNA48, RNA53, RNA44)
rownames(counts_matrix2) <- counts_matrix2$gene_id
counts_matrix2 <- counts_matrix2 %>%
  select(-gene_id)
counts_matrix2 <- as.matrix(counts_matrix2)  # Convert to numeric


samples <- c('OS0658','OS21014','OD21014', 'AD22', 'RNA35', 'RNA62', 'AD24','AD25','RNA41','RNA45','RNA47','RNA48','RNA53','RNA44')
label <- c('CTRL','CTRL','CTRL','POAG_HighConf','POAG_HighConf','POAG_HighConf','POAG_LowConf','POAG_LowConf','POAG_LowConf','POAG_LowConf','POAG_LowConf', 'POAG_LowConf', 'POAG_LowConf', 'POAG_LowConf')
labels <- data.frame(samples, label)
rownames(labels) <- labels$samples
labels <- labels %>%
  select(-samples)

 ## Create a DESeqDataSet object
dds2 <- DESeqDataSetFromMatrix(countData = counts_matrix2,
                              colData = labels,
                              design = ~ label)  # 'condition' should be a column in sample_info

# Filter low-count genes (optional)
#dds <- dds[rowSums(counts(dds)) > 10, ]

## Run DESeq analysis
dds2 <- DESeq(dds2)
res2 <- results(dds2)

```

```{r}
vsd_14 <- vst(dds2, blind = FALSE)

plotPCA(vsd_14, intgroup = "label") +
  theme_minimal() +
  ggtitle("PCA of RNA-seq Data: 14 Samples") +
  theme(plot.title = element_text(hjust = 0.5)) +
  ylim(-30, 15)
```


## Comparison

#### Identify overlap in differentially expressed genes. 

```{r}
#overlap <- intersect(all_genes_6, all_genes_11)
#length(overlap)

# Overlap with stricter p-value
overlap_adj <- intersect(all_genes_6_adj, all_genes_11_adj)
length(overlap_adj)
```


#### Jaccard similarity
```{r}
# Calculate the intersection and union of the sets
intersection <- length(overlap_adj)
union <- length(union(all_genes_6_adj, all_genes_11_adj))

# Calculate Jaccard Similarity
jaccard_similarity <- intersection / union
jaccard_similarity

```


From here, change significance level for genes in the contaminated subset (make stricter) to yield fewer genes. 
Change to 5e-20. Maybe want to find p-value that maximizes Jaccard similarity? Want to determine if the same genes shown significant in the high-confidence set is also present at high significant in the contaminated set, just maybe buried under contamination/noise. 


Use p-value of 5e-20 to extract significant genes, then calculate Jaccard similarity again:
```{r}
p_val <- 5e-20

# Extract significant genes
sig_genes_11_adj2 <- subset(res, pvalue < p_val & abs(log2FoldChange) > 1)
all_genes_11_adj2 <- rownames(sig_genes_11_adj2)

# Get overlap
overlap_adj2 <- intersect(all_genes_6_adj, all_genes_11_adj2)
length(overlap_adj2)

# Get Jaccard similarity
intersection <- length(overlap_adj2)
union <- length(union(all_genes_6_adj, all_genes_11_adj2))
jaccard_similarity <- intersection / union
jaccard_similarity
```


```{r}
p_value_list = c(5e-6, 5e-8, 5e-10, 5e-12, 5e-14, 5e-16, 5e-18, 5e-20, 5e-22, 5e-24, 5e-26)
jacc <- data.frame('p_val'=numeric(), 'length_overlap'=numeric(), 'jaccard_sim'=numeric())

for (p in p_value_list){
  sig_genes_11_adjj <- subset(res, pvalue < p & abs(log2FoldChange) > 1)
  all_genes_11_adjj <- rownames(sig_genes_11_adjj)
  
  overlap_adjj <- intersect(all_genes_6_adj, all_genes_11_adjj)
  
  intersection <- length(overlap_adjj)
  union <- length(union(all_genes_6_adj, all_genes_11_adjj))
  jaccard_similarity <- intersection / union
  
  new <- data.frame('p_val' = p, 'length_overlap'=union, 'jaccard_sim'=jaccard_similarity)
  jacc <- rbind(jacc, new)
}
```

```{r}
jacc$p_val <- factor(jacc$p_val)
ggplot(jacc, aes(x=p_val, y=jaccard_sim, size=length_overlap))+
  geom_point()+
  theme_minimal()+
  labs(title='Jaccard Similarity versus P-value', x='P-value', y='Jaccard Similarity')+
  theme(plot.title = element_text(hjust = 0.5))
```





#### Make table of overlapping genes
```{r}
df_res6 <- data.frame(res_6)
df_res11 <- data.frame(res)

# Convert row names to a column
df_res6$GENE <- rownames(df_res6)
df_res11$GENE <- rownames(df_res11)

# Merge
df_overlap <- merge(df_res6, df_res11, by = "GENE", all = FALSE)  # Inner join but doesn't matter cause same in both

# Now filter so that it's only the 32 overlapping genes
df_overlap <- df_overlap[df_overlap$GENE %in% overlap_adj, ]

df_overlap <- df_overlap %>%
  dplyr::rename(baseMean.6 = baseMean.x, log2FoldChange.6 = log2FoldChange.x, lfcSE.6 = lfcSE.x, stat.6 = stat.x, pvalue.6 = pvalue.x,
         padj.6 = padj.x, baseMean.11 = baseMean.y, log2FoldChange.11 = log2FoldChange.y, lfcSE.11 = lfcSE.y, stat.11 = stat.y,
         pvalue.11 = pvalue.y, padj.11 = padj.y)
```

```{r}
# Write to csv
#write.csv(df_overlap, 'GeneOverlap.csv', row.names = F)
```

```{r}
# Get most significant genes (most significant in 6-sample)

df_overlap2 = df_overlap %>%
  dplyr::rename(EnsemblGeneID = GENE) %>%
  arrange(pvalue.6) %>%
  mutate(DirectionOfExpression = ifelse(log2FoldChange.6 > 0, 'Upregulated', 'Downregulated')) %>%
  select(EnsemblGeneID, DirectionOfExpression) %>%
  head(10)


library(knitr)
library(dplyr)
library(kableExtra)

kable(df_overlap2, caption = "Top 10 Differentially Expressed Genes Across High- and Low-Confidence Samples") %>%
  column_spec(1, width = "6cm", extra_css = "padding-left: 20px;") %>%
  column_spec(2, width = "6cm", extra_css = "padding-left: 20px;") %>%
  row_spec(0, extra_css = "padding-left: 20px; text-align: left;")


```




## Functional Enrichment Analysis

```{r}
#install.packages(c("httr", "jsonlite", "curl"))
#install.packages("gprofiler2")
library(gprofiler2)
```

```{r}
# Run functional enrichment analysis
gost_results <- gost(query = overlap_adj, 
                     organism = "hsapiens",  # Human genes
                     sources = c("GO:BP", "GO:MF", "GO:CC", "KEGG", "REACTOME"), 
                     significant = TRUE)

# Print results
head(gost_results$result)
```

```{r}
# Extract top 10 enriched terms
top_terms <- gost_results$result[1:10,]

ggplot(top_terms, aes(x=reorder(term_name, -p_value), y=-log10(p_value))) +
  geom_bar(stat="identity", fill="steelblue") +
  coord_flip() +
  labs(title="Top Enriched Terms", x="Pathway/GO Term", y="-log10(p-value)") +
  theme_minimal()

```

```{r}
# Interactive manhattan plot
gostplot(gost_results, capped = TRUE, interactive = TRUE)
```




