---
title: "DE Full Report"
author: "Sam Remmey, James Siegener, Jacob Turner"
date: "May 2, 2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("BiocManager")  # If not already installed
#BiocManager::install("DESeq2")
#install.packages(c("httr", "jsonlite", "curl"))
#install.packages("gprofiler2")

library(tidyverse)
library(DESeq2)
library(ggplot2)
library(dplyr)
library(pheatmap)
library(ggrepel)
library(tibble)
library(knitr)
library(kableExtra)
library(gprofiler2)
```

## Differential Expression Analysis

We've conducted DE analyses on both the high confidence 6-sample subset (3 POAG, 3 cadaver) as well as an 11-sample subset we formed by combining the 3 high confidence cadaver control samples with the 8 low confidence POAG samples. This report contains results of both series of analyses as well as a list of extracted genes that were found to be differentially expressed in both subsets, followed by functional enrichment analysis conducted on this extracted set of genes. 


Global variables:
```{r}
# Threshold for significance: p-value of 5e-10 is commonly used
p_val <- 5e-10
```


### DE: 6 samples

```{r}
gene_6 <- readxl::read_excel("gene_count_6.xls")

counts_6 <- gene_6 %>% select(gene_id, OS0658, OS21014, OD21014, AD22, RNA35, RNA62)

counts_6 <- counts_6 %>% column_to_rownames(var = "gene_id")

counts_6 <- as.matrix(counts_6)

mode(counts_6) <- "numeric"

sample_info_6 <- data.frame(
    row.names = colnames(counts),   # Sample names
    condition = c("control", "control", "control", "glaucoma", "glaucoma", "glaucoma")  # Group labels
)

sample_info_6$condition <- factor(sample_info_6$condition, levels = c("control", "glaucoma"))
```

```{r}
dds_6 <- DESeqDataSetFromMatrix(countData = counts_6,
                              colData = sample_info_6,
                              design = ~ condition)

# Filter out low-expression genes
# dds_6 <- dds_6[rowSums(counts_6(dds_6)) > 10, ]

# Run DESeq2 normalization and differential expression analysis
dds_6 <- DESeq(dds_6)

# Extract results (glaucoma vs. control)
res_6 <- results(dds_6, contrast = c("condition", "glaucoma", "control"))

# View summary
summary(res_6)
```

```{r}
## Extract significant genes
sig_genes_6_adj <- subset(res_6, pvalue < p_val & abs(log2FoldChange) > 1)
all_genes_6_adj <- rownames(sig_genes_6_adj)
length(all_genes_6_adj)
```


```{r}
logfc_thresh <- 1  # threshold for large fold change

# Prepare data
res_df_6 <- as.data.frame(res_6) %>%
  rownames_to_column(var = "gene") %>%
  drop_na() %>%
  mutate(
    log2FoldChange = as.numeric(log2FoldChange),
    padj = as.numeric(padj),
    pvalue = as.numeric(pvalue),
    neg_log10_pvalue = -log10(pvalue),
    sig = case_when(
      pvalue < p_val ~ "Significant",  # using pvalue rather than padj because our manually set p_val is very strict
      TRUE ~ "Not Significant"))

# Label the top genes
topp_genes <- res_df_6 %>%
  filter(pvalue < p_val) %>%
  top_n(-10, pvalue)  # 10 genes with smallest p-values

# Volcano plot
ggplot(res_df_6, aes(x = log2FoldChange, y = neg_log10_pvalue, color = sig)) +
  geom_point(alpha = 0.7, size = 1.5) +
  scale_color_manual(values = c(
    "Significant" = "red",
    "Not Significant" = "grey60")) +
  geom_text_repel(
    data = topp_genes,
    aes(label = gene),
    size = 3,
    max.overlaps = 10,
    box.padding = 0.3,
    point.padding = 0.2,
    segment.color = 'grey50') +
  geom_hline(yintercept = -log10(p_val), linetype = "dashed", color = "gray40") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Volcano Plot of Differential Gene Expression: 6 Samples",
    x = "Log2 Fold Change",
    y = "-Log10 P-value",
    color = "Significance:") +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "top")

```


```{r}
# Ensure top genes match the volcano plot
topp_genes_6 <- topp_genes$gene  # pulled from volcano plot step

# Extract normalized expression matrix
vsd_6 <- vst(dds_6, blind = FALSE)  # DESeq2 object → variance-stabilized matrix
vsd_mat <- assay(vsd_6)

# Subset matrix for top genes
heatmap_data <- vsd_mat[topp_genes_6, ]

# Z-score normalize across samples for better visual contrast
heatmap_scaled <- t(scale(t(heatmap_data)))

# Build heatmap with labels and title
pheatmap(
  mat = heatmap_scaled,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
  main = "Heatmap of Top 10 Most Significant Genes: 6 Samples",
  fontsize_row = 10,
  fontsize_col = 10,
  border_color = NA,
  show_rownames = TRUE,
  show_colnames = TRUE,
  labels_row = topp_genes_6)

```

```{r}
# PCA

# Run PCA
pcaData <- plotPCA(vsd_6, intgroup = "condition", returnData = TRUE)

# Calculate % variance explained by PC1 and PC2
percentVar <- round(100 * attr(pcaData, "percentVar"))

# Plot
ggplot(pcaData, aes(x = PC1, y = PC2, color = condition)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  ggtitle("PCA of RNA-seq Expression (6 Samples)") +
  theme_minimal(base_size = 14) +
  scale_color_manual(values = c("glaucoma" = "firebrick3", "control" = "navy")) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.title = element_blank())

```


### DE: 11 samples

```{r}
gene_16 <- data.frame(readxl::read_excel("gene_count_16.xls"))

counts_matrix <- gene_16 %>%
  select(gene_id, OS0658, OS21014, OD21014, AD24, AD25, RNA41, RNA45, RNA47, RNA48, RNA53, RNA44)
rownames(counts_matrix) <- counts_matrix$gene_id
counts_matrix <- counts_matrix %>%
  select(-gene_id)
counts_matrix <- as.matrix(counts_matrix)  # Convert to numeric


samples <- c('OS0658','OS21014','OD21014','AD24','AD25','RNA41','RNA45','RNA47','RNA48','RNA53','RNA44')
label <- c('control','control','control','glaucoma','glaucoma','glaucoma','glaucoma','glaucoma','glaucoma','glaucoma','glaucoma')
labels <- data.frame(samples, label)
rownames(labels) <- labels$samples
labels <- labels %>%
  select(-samples)
```

```{r}
 ## Create a DESeqDataSet object
dds <- DESeqDataSetFromMatrix(countData = counts_matrix,
                              colData = labels,
                              design = ~ label)

# Filter low-count genes (optional)
#dds <- dds[rowSums(counts(dds)) > 10, ]

## Run DESeq analysis
dds <- DESeq(dds)
res <- results(dds)
```

```{r}
# p values
res <- res[order(res$pvalue), ]
head(res)

# Extract significant genes
sig_genes_11_adj <- subset(res, pvalue < p_val & abs(log2FoldChange) > 1)
all_genes_11_adj <- rownames(sig_genes_11_adj)
```

```{r}
# Compare number of significant genes from 11-sample vs 6-sample
length(all_genes_11_adj)
length(all_genes_6_adj)
```

Note the huge difference in number of significant genes here. This result suggests expression in the 11-sample subset may contain a lot of noise. 


```{r}
logfc_thresh <- 1  # threshold for large fold change

# Prepare data
res_df_11 <- as.data.frame(res) %>%
  rownames_to_column(var = "gene") %>%
  drop_na() %>%
  mutate(
    log2FoldChange = as.numeric(log2FoldChange),
    padj = as.numeric(padj),
    pvalue = as.numeric(pvalue),
    neg_log10_pvalue = -log10(pvalue),
    sig = case_when(
      pvalue < p_val ~ "Significant",  # using pvalue rather than padj because our manually set p_val is very strict
      TRUE ~ "Not Significant"))

# Label the top genes
topp_genes11 <- res_df_11 %>%
  filter(pvalue < p_val) %>%
  top_n(-10, pvalue)  # 10 genes with smallest p-values

# Volcano plot
ggplot(res_df_11, aes(x = log2FoldChange, y = neg_log10_pvalue, color = sig)) +
  geom_point(alpha = 0.7, size = 1.5) +
  scale_color_manual(values = c(
    "Significant" = "red",
    "Not Significant" = "grey60")) +
  geom_text_repel(
    data = topp_genes11,
    aes(label = gene),
    size = 3,
    max.overlaps = 20,
    box.padding = 0.3,
    point.padding = 0.2,
    segment.color = 'grey50') +
  geom_hline(yintercept = -log10(p_val), linetype = "dashed", color = "gray40") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Volcano Plot of Differential Gene Expression: 11 Samples",
    x = "Log2 Fold Change",
    y = "-Log10 P-value",
    color = "Significance:") +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "top")


```

```{r}
# Ensure top genes match the volcano plot
topp_genes_11 <- topp_genes11$gene  # pulled from volcano plot step

# Extract normalized expression matrix
vsd_11 <- vst(dds, blind = FALSE)  # DESeq2 object → variance-stabilized matrix
vsd_mat11 <- assay(vsd_11)

# Subset matrix for top genes
heatmap_data11 <- vsd_mat11[topp_genes_11, ]

# Z-score normalize across samples for better visual contrast
heatmap_scaled11 <- t(scale(t(heatmap_data11)))

# Build heatmap with labels and title
pheatmap(
  mat = heatmap_scaled11,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
  main = "Heatmap of Top 10 Most Significant Genes: 11 Samples",
  fontsize_row = 10,
  fontsize_col = 10,
  border_color = NA,
  show_rownames = TRUE,
  show_colnames = TRUE,
  labels_row = topp_genes_11)

```


```{r}
# PCA

# Run PCA
pcaData11 <- plotPCA(vsd_11, intgroup = "label", returnData = TRUE)

# Calculate % variance explained by PC1 and PC2
percentVar <- round(100 * attr(pcaData11, "percentVar"))

# Plot
ggplot(pcaData11, aes(x = PC1, y = PC2, color = label)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  ggtitle("PCA of RNA-seq Expression (11 Samples)") +
  theme_minimal(base_size = 14) +
  scale_color_manual(values = c("glaucoma" = "firebrick3", "control" = "navy")) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.title = element_blank())
```


PCA again: this time add back the 3 high confidence POAG samples and run PCA on that.
```{r}
gene_16 <- data.frame(readxl::read_excel("gene_count_16.xls"))

counts_matrix2 <- gene_16 %>%
  select(gene_id, OS0658, OS21014, OD21014, AD22, RNA35, RNA62, AD24, AD25, RNA41, RNA45, RNA47, RNA48, RNA53, RNA44)
rownames(counts_matrix2) <- counts_matrix2$gene_id
counts_matrix2 <- counts_matrix2 %>%
  select(-gene_id)
counts_matrix2 <- as.matrix(counts_matrix2)  # Convert to numeric


samples <- c('OS0658','OS21014','OD21014', 'AD22', 'RNA35', 'RNA62', 'AD24','AD25','RNA41','RNA45','RNA47','RNA48','RNA53','RNA44')
label <- c('control','control','control','POAG_HighConf','POAG_HighConf','POAG_HighConf','POAG_LowConf','POAG_LowConf','POAG_LowConf','POAG_LowConf','POAG_LowConf', 'POAG_LowConf', 'POAG_LowConf', 'POAG_LowConf')
labels <- data.frame(samples, label)
rownames(labels) <- labels$samples
labels <- labels %>%
  select(-samples)

 ## Create a DESeqDataSet object
dds2 <- DESeqDataSetFromMatrix(countData = counts_matrix2,
                              colData = labels,
                              design = ~ label)

# Filter low-count genes (optional)
#dds <- dds[rowSums(counts(dds)) > 10, ]

## Run DESeq analysis
dds2 <- DESeq(dds2)
res2 <- results(dds2)

```

```{r}
vsd_14 <- vst(dds2, blind = FALSE)

plotPCA(vsd_14, intgroup = "label") +
  theme_minimal() +
  ggtitle("PCA of RNA-seq Data: 14 Samples") +
  theme(plot.title = element_text(hjust = 0.5)) +
  ylim(-30, 15)
```


## Comparison

#### Identify overlap in differentially expressed genes. 

```{r}
# Overlap with strict p-value
overlap_adj <- intersect(all_genes_6_adj, all_genes_11_adj)
length(overlap_adj)
```


#### Jaccard similarity
```{r}
# Calculate the intersection and union of the sets
intersection <- length(overlap_adj)
union <- length(union(all_genes_6_adj, all_genes_11_adj))

# Calculate Jaccard Similarity
jaccard_similarity <- intersection / union
jaccard_similarity

```


From here, we will change the significance level for genes in the contaminated subset (make stricter) to yield fewer genes. 

Maybe we want to find the p-value that maximizes Jaccard similarity? We want to determine if the same genes that are significant in the high-confidence set are also present at high significance in the contaminated set, just maybe buried under contamination/noise. 


Use p-value of 5e-20 to extract significant genes, then calculate Jaccard similarity again:
```{r}
p_val <- 5e-20

# Extract significant genes
sig_genes_11_adj2 <- subset(res, pvalue < p_val & abs(log2FoldChange) > 1)
all_genes_11_adj2 <- rownames(sig_genes_11_adj2)

# Get overlap
overlap_adj2 <- intersect(all_genes_6_adj, all_genes_11_adj2)
length(overlap_adj2)

# Get Jaccard similarity
intersection <- length(overlap_adj2)
union <- length(union(all_genes_6_adj, all_genes_11_adj2))
jaccard_similarity <- intersection / union
jaccard_similarity
```


Now we will iterate over several p-value thresholds. 
```{r}
p_value_list = c(5e-6, 5e-8, 5e-10, 5e-12, 5e-14, 5e-16, 5e-18, 5e-20, 5e-22, 5e-24, 5e-26)
jacc <- data.frame('p_val'=numeric(), 'length_overlap'=numeric(), 'jaccard_sim'=numeric())

for (p in p_value_list){
  sig_genes_11_adjj <- subset(res, pvalue < p & abs(log2FoldChange) > 1)
  all_genes_11_adjj <- rownames(sig_genes_11_adjj)
  
  overlap_adjj <- intersect(all_genes_6_adj, all_genes_11_adjj)
  
  intersection <- length(overlap_adjj)
  union <- length(union(all_genes_6_adj, all_genes_11_adjj))
  jaccard_similarity <- intersection / union
  
  new <- data.frame('p_val' = p, 'length_overlap'=union, 'jaccard_sim'=jaccard_similarity)
  jacc <- rbind(jacc, new)
}
```

```{r}
jacc$p_val <- factor(jacc$p_val)
ggplot(jacc, aes(x=p_val, y=jaccard_sim, size=length_overlap))+
  geom_point()+
  theme_minimal()+
  labs(title='Jaccard Similarity versus P-value', x='P-value', y='Jaccard Similarity')+
  theme(plot.title = element_text(hjust = 0.5))
```


#### Make table of overlapping genes
```{r}
df_res6 <- data.frame(res_6)
df_res11 <- data.frame(res)

# Convert row names to a column
df_res6$GENE <- rownames(df_res6)
df_res11$GENE <- rownames(df_res11)

# Merge
df_overlap <- merge(df_res6, df_res11, by = "GENE", all = FALSE)  # Inner join but doesn't matter cause same in both

# Now filter so that it's only the 32 overlapping genes
df_overlap <- df_overlap[df_overlap$GENE %in% overlap_adj, ]

df_overlap <- df_overlap %>%
  dplyr::rename(baseMean.6 = baseMean.x, log2FoldChange.6 = log2FoldChange.x, lfcSE.6 = lfcSE.x, stat.6 = stat.x, pvalue.6 = pvalue.x,
         padj.6 = padj.x, baseMean.11 = baseMean.y, log2FoldChange.11 = log2FoldChange.y, lfcSE.11 = lfcSE.y, stat.11 = stat.y,
         pvalue.11 = pvalue.y, padj.11 = padj.y)
```


```{r}
# Write to csv
#write.csv(df_overlap, 'GeneOverlap.csv', row.names = F)
```

```{r}
# Get most significant genes (most significant in 6-sample)

df_overlap2 = df_overlap %>%
  dplyr::rename(EnsemblGeneID = GENE) %>%
  arrange(pvalue.6) %>%
  mutate(DirectionOfExpression = ifelse(log2FoldChange.6 > 0, 'Upregulated', 'Downregulated')) %>%
  select(EnsemblGeneID, DirectionOfExpression) %>%
  head(10)

kable(df_overlap2, caption = "Top 10 Differentially Expressed Genes Across High- and Low-Confidence Samples") %>%
  column_spec(1, width = "6cm", extra_css = "padding-left: 20px;") %>%
  column_spec(2, width = "6cm", extra_css = "padding-left: 20px;") %>%
  row_spec(0, extra_css = "padding-left: 20px; text-align: left;")


```


Another volcano plot:
```{r}
filtered_6 <- df_res6 %>% filter(GENE %in% overlap_adj)
filtered_11 <- df_res11 %>% filter(GENE %in% overlap_adj)

filtered_6$confidence <- "High (6 Samples)"
filtered_11$confidence <- "Low (11 Samples)"


overlap_data <- bind_rows(filtered_6, filtered_11)


overlap_data <- overlap_data %>%
  group_by(GENE) %>%
  filter(n() == 2) %>%  
  filter(all(log2FoldChange > 0) | all(log2FoldChange < 0)) %>% 
  ungroup()
```

```{r}
ggplot(overlap_data, aes(x = log2FoldChange, y = -log10(pvalue), color = as.factor(confidence))) +
  geom_point(alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black", linewidth = 1) + 
  scale_color_manual(values = c("High (6 Samples)" = "blue", "Low (11 Samples)" = "red")) +
  theme_minimal() +
  xlab("Log2 Fold Change") +
  ylab("-Log10 P-Value") +
  ggtitle("Volcano Plot of Differentially Expressed Genes") +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(color = "Confidence Level")
```
The above plot contains a point for each gene that was found to be significantly differentially expressed in both the high- and low-confidence subsets. There are 32 of these genes in total. 


#### Correlating DE results

```{r}
dat <- read.csv('GeneOverlap.csv')

# Overlap - significant in both
cor(dat$baseMean.6, dat$baseMean.11) # 0.99

# All genes
cor(df_res6$baseMean, df_res11$baseMean) # 0.01

# Only genes significantly differentially expressed in 6 samples
only6_6 <- df_res6[rownames(df_res6) %in% all_genes_6_adj, ]
only6_11 <- df_res11[rownames(df_res6) %in% all_genes_6_adj, ]
cor(only6_6$baseMean, only6_11$baseMean) # 0.15
```


```{r}
# Genes significant across 6 samples and 11 samples
ggplot(dat, aes(x=baseMean.6, y=baseMean.11))+geom_point()+theme_minimal()
```

## Functional Enrichment Analysis

```{r}
#install.packages(c("httr", "jsonlite", "curl"))
#install.packages("gprofiler2")
library(gprofiler2)
```

```{r}
# Run functional enrichment analysis
gost_results <- gost(query = overlap_adj, # 32 overlapping genes
                     organism = "hsapiens",  # Human genes
                     sources = c("GO:BP", "GO:MF", "GO:CC", "KEGG", "REACTOME"), 
                     significant = TRUE)

# Print results
head(gost_results$result)
```

```{r}
# Extract top 10 enriched terms
top_terms <- gost_results$result[1:10,]

ggplot(top_terms, aes(x=reorder(term_name, -p_value), y=-log10(p_value))) +
  geom_bar(stat="identity", fill="steelblue") +
  coord_flip() +
  labs(title="Top Enriched Terms: Overlap in High- and Low-Confidence Samples", x="Pathway/GO Term", y="-log10(p-value)") +
  theme_minimal()

```

```{r}
# Interactive manhattan plot
gostplot(gost_results, capped = TRUE, interactive = TRUE)
```


Now again with only genes significant in high confidence (6 sample) subset. 
```{r}
# Run functional enrichment analysis
gost_results2 <- gost(query = all_genes_6_adj, # 64 genes
                     organism = "hsapiens",  # Human genes
                     sources = c("GO:BP", "GO:MF", "GO:CC", "KEGG", "REACTOME"), 
                     significant = TRUE)

# Print results
head(gost_results2$result)
```

```{r}
# Extract top 10 enriched terms
top_terms2 <- gost_results2$result[1:10,]

ggplot(top_terms2, aes(x=reorder(term_name, -p_value), y=-log10(p_value))) +
  geom_bar(stat="identity", fill="steelblue") +
  coord_flip() +
  labs(title="Top Enriched Terms: High Confidence Samples Only", x="Pathway/GO Term", y="-log10(p-value)") +
  theme_minimal()
```

```{r}
# Interactive manhattan plot
gostplot(gost_results2, capped = TRUE, interactive = TRUE)
```


## GWAS Comparison
```{r}
library(org.Hs.eg.db)
POAG_associated_genes <- read.delim("EFO_0004190_associations_export.tsv")
#Rename significant our gene data sets for clarity 
ensemble_genes_6 <- all_genes_6_adj
ensemble_genes_11 <- all_genes_11_adj

#Convert ensemble gene codes to actual gene names
gene_names_6 <- AnnotationDbi::select(org.Hs.eg.db, keys = ensemble_genes_6, columns = "SYMBOL", keytype = "ENSEMBL")
gene_names_11 <- AnnotationDbi::select(org.Hs.eg.db, keys = ensemble_genes_11, columns = "SYMBOL", keytype = "ENSEMBL")

#Get the unique genes from the GWAS study that are significantly differentially expressed at the significance level we are using (5e-8)
significant_GWAS_genes <- POAG_associated_genes[POAG_associated_genes$pValue <= 5e-8, ]
split_genes <- significant_GWAS_genes %>%
  separate_rows(mappedGenes, sep= ",")
split_genes$mappedGenes <- trimws(split_genes$mappedGenes)
unique_genes <- unique(split_genes$mappedGenes)

#Find the overlap between the genes we identified and the genes from the GWAS study
common_genes_6 <- intersect(unique_genes, gene_names_6$SYMBOL)
common_genes_11 <- intersect(unique_genes, gene_names_11$SYMBOL)
common_genes <- intersect(common_genes_11, common_genes_6)
```



