---
title: 'PCA with Harmonization'
author: "Sam Remmey, James Siegener, Jacob Turner"
date: "May 2, 2025"
output: html_document
---

### Background 

"Harmonization" here refers to the process of standardizing or normalizing gene expression data from different sources to make them comparable. In our case, we were hoping to utilize harmonization techniques to mitigate potential batch effects and examine PCA results after applying this additional normalization step. Given the level of contamination present in the data, we believe that more rigorous analytical techniques would be needed to account more fully for the added noise. An extension of this project could include an investigation into these aforementioned analytical techniques. 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("BiocManager")  # If not already installed
#BiocManager::install("DESeq2")

library(tidyverse)
library(DESeq2)
library(ggplot2)
library(pheatmap)
library(limma)
```

Global variables:
```{r}
p_val <- 5e-10
```


### Pre-processing: harmonization

```{r}
#BiocManager::install("sva")
#library(sva)
```

```{r}
gene_16 <- data.frame(readxl::read_excel("gene_count_16.xls"))

counts_matrix2 <- gene_16 %>%
  select(gene_id, OS0658, OS21014, OD21014, AD22, RNA35, RNA62, AD24, AD25, RNA41, RNA45, RNA47, RNA48, RNA53, RNA44)
rownames(counts_matrix2) <- counts_matrix2$gene_id
counts_matrix2 <- counts_matrix2 %>%
  select(-gene_id)
counts_matrix2 <- as.matrix(counts_matrix2)  # Convert to numeric


samples <- c('OS0658','OS21014','OD21014', 'AD22', 'RNA35', 'RNA62', 'AD24','AD25','RNA41','RNA45','RNA47','RNA48','RNA53','RNA44')
label <- c('CTRL','CTRL','CTRL','POAG_HighConf','POAG_HighConf','POAG_HighConf','POAG_LowConf','POAG_LowConf','POAG_LowConf','POAG_LowConf','POAG_LowConf', 'POAG_LowConf', 'POAG_LowConf', 'POAG_LowConf')
labels <- data.frame(samples, label)
rownames(labels) <- labels$samples
labels <- labels %>%
  select(-samples)


 ## Create a DESeqDataSet object
dds2 <- DESeqDataSetFromMatrix(countData = counts_matrix2,
                              colData = labels,
                              design = ~ label) 
```


Apply vst()
```{r}
# Apply VST
vsd_14 <- vst(dds2, blind = FALSE)
```


Harmonization with limma::removeBatchEffect()
```{r}
# Define batch vector: 1 = Contaminated (POAG_LowConf), 0 = Not Contaminated (CTRL, POAG_HighConf)
batch <- ifelse(labels$label == "POAG_LowConf", 1, 0)

# Apply batch correction to the VST-transformed data
assay(vsd_14) <- removeBatchEffect(assay(vsd_14), batch = batch)

```


Visualize PCA plot after harmonization
```{r}
plotPCA(vsd_14, intgroup = "label") +
  theme_minimal() +
  ggtitle("PCA of RNA-seq Data After Harmonization") +
  theme(plot.title = element_text(hjust = 0.5))

```







